### 2016-06-24 publications/Journal/3-features.tex
Re-add publication folder + draft journal paper.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

Historically, dxex is based on PythonPDEVS~\cite{PythonPDEVS}.
Python is a good language for prototypes, but performance has proven insufficient to compete with other simulation kernels~\cite{MasterThesis}.
Dxex is a C++11-based implementation of PythonPDEVS, but implements only a subset of PythonPDEVS, while making some of its own additions.
So while the feature set is not too comparable, the architectural design, core simulation algorithm, and optimizations, are highly similar.

We will not make a detailed comparison with PythonPDEVS here, but only mention some supported features.
Dxex supports, similarly to PythonPDEVS, the following features: direct connection~\cite{SymbolicFlattening}, \textsf{Dynamic Structure DEVS}~\cite{DSDEVS}, termination conditions, and a modular tracing and scheduling framework~\cite{PythonPDEVS}.
But whereas PythonPDEVS only supports optimistic synchronization, dxex support multiple synchronization protocols (though only in parallel).
This is in line with the design principle of PythonPDEVS: allow users to pass performance hints to the simulation kernel.
In our case, a user can pass the simulation kernel the synchronization protocol to use for this model.
Our implementation in C++11 also allows for optimizations which were plainly impossible in an interpreted language.

Since there is no universal \textsf{DEVS} model standard, dxex models are incompatible with PythonPDEVS and vice versa.
This is due to dxex models being grafted on C++11, whereas PythonPDEVS models are grafted on Python.

@@

\subsection{Synchronization protocols}
We previously explained the existence of different synchronization protocols exist, each optimized for a specific kind of model.
As no single synchronization protocol is ideal for all models, a general purpose simulation tool should support multiple protocols.
Currently, most parallel simulation tools choose only a single synchronization protocol due to the inherent differences between protocols.
An uninformed choice on which one to implement is insufficient, as performance will likely be bad.
We argue that a real general purpose simulation tool should support sequential, conservative, and optimistic synchronization, as is the case for dxex.

@@

\begin{figure}
    \includegraphics[width=\columnwidth]{fig/cores_class_diagram.eps}
	\caption{Dxex kernel design.}
	\label{fig:class_diagram}
\end{figure}

@@

Reading and writing to shared memory is done through the use of the new C++11 synchronization primitives.
Whereas this was also possible in previous versions of the C++ standard, by falling back to non-portable C functions, it was not a part of the C++ language standard.
C++11 further allows us to make the implementation portable, as well as more efficient: the compiler might know of optimizations specific to atomic variables.

@@

The main problem we encountered in our implementation is the aggressive use of memory.
Frequent memory allocation and deallocation caused significant overheads, certainly when multiple threads do so concurrently.
This made us switch to the use of thread-local (using \texttt{tcmalloc}) memory pools.
Again, we made use of specific new features of C++11, that were not available in Python, or even previous versions of the C++ language standard.

\subsection{Transparency}
We define simulation kernel transparency as having a single model, which always can be executed on each supported synchronization kernel, without any modifications.
User should thus only provide one model, implemented in C++11, which can be either using sequential execution, using conservative synchronization, or using optimistic synchronization.
Switching between simulation kernels is as simple as altering the simulation termination time.
The exception is conservative synchronization, where a lookahead function is required, which is not used in other synchronization kernels.
Two options are possible: either a lookahead function must always be provided, even when it is not required and possibly not used, or we use a default lookahead function if none is defined.

@@

>>>>

\newcommand{\plotfraction}{0.85}
Historically, dxex is based on PythonPDEVS~\cite{PythonPDEVS}.
Python is a good language for prototypes, but performance has proven insufficient to compete with other simulation kernels~\cite{MasterThesis}.
Dxex is a C++14-based implementation of PythonPDEVS, but implements only a subset of PythonPDEVS, while making some of its own additions.
So while the feature set is not too comparable, the architectural design, core simulation algorithm, and optimizations, are highly similar.

We will not make a detailed comparison with PythonPDEVS here, but only mention some supported features.
Dxex supports, similarly to PythonPDEVS, the following features: direct connection~\cite{SymbolicFlattening}, \textsf{Dynamic Structure DEVS}~\cite{DSDEVS}, termination conditions, and a modular tracing and scheduling framework~\cite{PythonPDEVS}.
But whereas PythonPDEVS only supports optimistic synchronization, dxex support multiple synchronization protocols (though only in parallel).
This is in line with the design principle of PythonPDEVS: allow users to pass performance hints to the simulation kernel.
In our case, a user can pass the simulation kernel the synchronization protocol to use for this model.
Our implementation in C++14 also allows for optimizations which were plainly impossible in an interpreted language.

Since there is no universal \textsf{DEVS} model standard, dxex models are incompatible with PythonPDEVS and vice versa.
This is due to dxex models being grafted on C++14, whereas PythonPDEVS models are grafted on Python.

@@

\subsection{Synchronization protocols}
We previously explained the existence of different synchronization protocols, each optimized for a specific kind of model.
As no single synchronization protocol is ideal for all models, a general purpose simulation tool should support multiple protocols.
Currently, most parallel simulation tools choose only a single synchronization protocol due to the inherent differences between protocols.
An uninformed choice on which one to implement is insufficient, as performance will likely be bad.
We argue that a real general purpose simulation tool should support sequential, conservative, and optimistic synchronization, as is the case for dxex.

@@

\begin{figure}
    \includegraphics[width=\columnwidth]{fig/cores_class_diagram.eps}
	\caption{Dxex design.}
	\label{fig:class_diagram}
\end{figure}

@@

Reading and writing to shared memory is done through the use of the new C++14 synchronization primitives.
Whereas this was also possible in previous versions of the C++ standard, by falling back to non-portable C functions, it was not a part of the C++ language standard.
C++14 further allows us to make the implementation portable, as well as more efficient: the compiler might know of optimizations specific to atomic variables.

@@

The main problem we encountered in our implementation is the aggressive use of memory.
Frequent memory allocation and deallocation caused significant overheads, certainly when multiple threads do so concurrently.
This made us switch to the use of thread-local (using \texttt{tcmalloc}) memory pools.
Again, we made use of specific new features of C++14, that were not available in Python, or even previous versions of the C++ language standard.

\subsection{Transparency}
We define simulation kernel transparency as having a single model, which always can be executed on each supported synchronization kernel, without any modifications.
User should thus only provide one model, implemented in C++14, which can be either using sequential execution, using conservative synchronization, or using optimistic synchronization.
Switching between simulation kernels is as simple as altering the simulation termination time.
The exception is conservative synchronization, where a lookahead function is required, which is not used in other synchronization kernels.
Two options are possible: either a lookahead function must always be provided, even when it is not required and possibly not used, or we use a default lookahead function if none is defined.

@@

\subsection{Features}
In this section we highlight some of the extra features dxex offers to modellers.
\subsubsection{Tracing}
Like PythonPDEVS dxex has full tracing capabilities supporting a wide number of output formats : json, XML, cell, console, ... . The user can implement his own outputpolicy and use it in our modular tracing design. The tracing functionality itself is implemented asynchronous and thread-safe. By overriding the output functions in the Message or State classes the user can generate model specific output for use with an existing tracing output policy. Tracing can be completely disabled to maximize simulation performance should this be desired (e.g. benchmarks). 
% Talk about adevs?
\subsubsection{Memory Allocation}
As we will demonstrate in \ref{sec:4-subsec:overhead-pgraph:memory} memory allocation is a performance bottleneck in simulation with high event frequency, especially in parallel. Dxex has a modular allocation interface that separates the object creation in kernels and models from the actual implementation. This allows us by using a single compilation parameter to change the allocation strategy per kernel type. This interface allows instrumentation of memory usage. Not only useful to detect memory leaks should this be a concern, it can also be of use to a model author to trace event creation in large models. A model author can rely on the default implementation as the fastest common configuration verified by the benchmarks.
\subsubsection{Model Allocation}
A model author can specify which kernel a model should be allocated to, should such manual intervention be required. This is handled by the default model allocator. If no preference is given a simple striping scheme is followed. It should be obvious that this is not sufficient in most simulations. In \ref{sec:4-performance} we analyze allocation schemes and their performance impact. By overriding the default allocator a model author can tune the allocation scheme to maximize parallel speedup. This interface can be in future work also be used to employ graph algorithms for an automatic allocation scheme, for example one that avoids cycles in the resulting kernel dependency graph.
\subsubsection{Logging}
A high performance asynchronous buffered threaded logging module is used in dxex which offers a complete overview of the entire simulation. The logging functionality can be used in any model, it only requires linking to the dxex library. Logging statements are written to file with a granular level system. In case of a severe crash the logging system will ensure persistence of the log file itself up until the last statement. Logging can be completely disabled to remove the performance impact.
\subsubsection{Visualization and Statistics}
Tracing offers a black box view of a DEVS model in a simulation, with transitions, events and states recorded at each state change. Dxex supplements this optionally with more in depth insights into the actual simulation. The allocation of models across kernels is visualized at the end of a simulation, with graphical representation of message frequency. This allows a user to fine tune future allocation schemes and the interface used by the visualization tool can also be used for more advanced purposes such as dynamic tracing of the simulation.\\
Invaluable during the development of dxex were runtime statistics collected by a simulation kernel. These record any measurable event in the simulation such as : events created/destroyed, inter kernel event frequency, reverts, GVT calculations, lookahead and eot values and insights into the fairness between the kernels themselves. Using this information it is possible to detect an imbalance between the simulation rounds that a single kernel executes compared to dependent or influencing kernels, or observe how balanced inter kernel message traffic actually is during a simulation. The same information can, as in the visualization tool, be used programmatically to extend dxex with visualization software that shows the simulation from a white box perspective or allow activity tracking.
Figures \ref{fig:pholdtree_visualize_seq},\ref{fig:pholdtree_visualize_parDFS},\ref{fig:pholdtree_visualize_parBFS} give an example of the modular visualization output generated by dxex after simulation the PholdTree model. These figures demonstrate the effect of allocation visually, the high inter kernels event count will degrade performance, as we will demonstrate in 
\ref{PholdTreeallocation}. Dotted lines represent connections between models that did not see events exchanged, directed edges represents events with frequency as weight and color to distinguis inter versus intra kernel events.
% Add figure
\begin{figure}
	\center
	\includegraphics[width=\plotfraction\columnwidth, height=8cm, keepaspectratio]{fig/pholdtreed1n3t5000.eps}
	\caption{Visualization of PholdTree (d=1,n=3,t=5000) sequential simulation.}
	\label{fig:pholdtree_visualize_seq}
\end{figure}
\begin{figure}
	\center
	\includegraphics[width=\plotfraction\columnwidth,  height=8cm, keepaspectratio]{fig/pholdtreed1n3t5000c4BFS.eps}
	\caption{Visualization of PholdTree (d=1,n=3,t=5000) parallel simulation with breadth first allocation and 4 kernels.}
	\label{fig:pholdtree_visualize_parBFS}
\end{figure}
\begin{figure}
	\center
	\includegraphics[width=\plotfraction\columnwidth,  height=8cm, keepaspectratio]{fig/pholdtreed1n3t5000c4DFS.eps}
	\caption{Visualization of PholdTree (d=1,n=3,t=5000) parallel simulation with depth first allocation and 4 kernels.}
	\label{fig:pholdtree_visualize_parDFS}
\end{figure}

### 2016-06-24 publications/Journal/6-conclusions.tex
Re-add publication folder + draft journal paper.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

In this paper, we introduced DEVS-Ex-Machina (``dxex''), a new C++11-based \textsf{Parallel DEVS} simulation tool.
Our main contribution is the implementation of multiple synchronization protocols for parallel multicore simulation.
We have shown that there are indeed models which can be simulated significantly faster using either synchronization protocol.
Dxex allows the user to choose between either conservative or optimistic synchronization as simple as any other configuration option.
Notwithstanding this modularity, dxex achieves performance competitive to adevs, another very efficient \textsf{DEVS} simulation tool.
Performance is measured both in elapsed time, and memory usage.

Future work is possible in several directions.
Firstly, our implementation of optimistic synchronization should be more tolerant to low-memory situations.
In its current state, simulation will simply halt with an out-of-memory error.
Having simulation control, which can throttle the speed of nodes that use up too much memory, has been shown to work in these situations~\cite{FujimotoBook}.
Faster GVT implementations, such as those presented by~\cite{Fujimoto:1997:CGV:268403.268404} and~\cite{Bauer:2005:SND:1069810.1070159}, might further help to minimize this problem.
Secondly, the idea of activity can be implemented for our simulation kernels, making it possible to dynamically switch between conservative and optimistic synchronization when behavioural changes are detected.
Thirdly, activity algorithms, as already implemented by PythonPDEVS, can also be implemented in dxex, to determine how they influence simulation performance.

>>>>

In this paper, we introduced DEVS-Ex-Machina (``dxex''), a new C++14-based \textsf{Parallel DEVS} simulation tool.
Our main contribution is the implementation of multiple synchronization protocols for parallel multicore simulation.
We have shown that there are indeed models which can be simulated significantly faster using either synchronization protocol.
Dxex allows the user to choose between either conservative or optimistic synchronization as simple as any other configuration option.
Notwithstanding this modularity, dxex achieves performance competitive to adevs, another very efficient \textsf{DEVS} simulation tool.
Performance is measured both in elapsed time, and memory usage.
Our empirical analysis shows that allocation of models over kernels is critical to enable a parallel speedup. Furthermore we have shown when and why optimistic synchronization can outperform conservative and vice versa.

Future work is possible in several directions.
Firstly, our implementation of optimistic synchronization should be more tolerant to low-memory situations.
In its current state, simulation will simply halt with an out-of-memory error.
Having simulation control, which can throttle the speed of nodes that use up too much memory, has been shown to work in these situations~\cite{FujimotoBook}.
Faster GVT implementations, such as those presented by~\cite{Fujimoto:1997:CGV:268403.268404} and~\cite{Bauer:2005:SND:1069810.1070159}, might further help to minimize this problem.
Secondly, the idea of activity can be implemented for our simulation kernels, making it possible to dynamically switch between conservative and optimistic synchronization when behavioural changes are detected.
Thirdly, activity algorithms, as already implemented by PythonPDEVS, can also be implemented in dxex, to determine how they influence simulation performance.
Finally automatic allocation is possible by analysis of the connections between models. This information is already used in dxex to determine the blocking order for conservative synchronization and in the directconnect algorithm. A graph algorithm that distributes models while avoiding cycles in the resulting kernels topology could be used to maximize parallel speedup in either optimistic or conservative synchronization.

### 2016-06-24 publications/Journal/abstract.tex
Re-add publication folder + draft journal paper.
Signed-off-by: Ben Cardoen <bencardoen@gmail.com>

With the ever increasing complexity of simulation models, parallel simulation
becomes necessary to perform the simulation within reasonable time bounds.
The built-in parallelism of Parallel DEVS is often insufficient to tackle this problem on its own.
Several synchronization protocols have been proposed, each with their distinct advantages and disadvantages.
Due to the significantly different implementation of these protocols, most Parallel DEVS simulation tools are limited to only one such protocol.
In this paper, we present a Parallel DEVS simulator, grafted on C++11, but based on PythonPDEVS, which supports both conservative and optimistic synchronization.
We evaluate the performance gain obtained by choosing the most appropriate synchronization protocol.
Performance results are compared to adevs, in terms of CPU time and memory usage.

>>>>

With the ever increasing complexity of simulation models, parallel simulation
becomes necessary to perform the simulation within reasonable time bounds.
The built-in parallelism of Parallel DEVS is often insufficient to tackle this problem on its own.
Several synchronization protocols have been proposed, each with their distinct advantages and disadvantages.
Due to the significantly different implementation of these protocols, most Parallel DEVS simulation tools are limited to only one such protocol.
In this paper, we present a Parallel DEVS simulator, grafted on C++14, but based on PythonPDEVS, which supports both conservative and optimistic synchronization.
We evaluate the performance gain obtained by choosing the most appropriate synchronization protocol.
Performance results are compared to adevs, in terms of CPU time and memory usage.
